---
title: "Average Mathematics Achievement by US Counties From Year of 2009 - 2016"
author: "Lok H. Chau"
date: 2018-05-13T1:13:15-05:00
categories: ["R"]
tags: ["R Markdown", "plot", "regression"]
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r results = 'hide', echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(sp)
library(rgeos)
library(rgdal)
library(maptools)
library(leaflet)
library(scales)
setwd('C:/Users/LokHeng/Desktop/SEDA/')
naep <- read.csv('County/SEDA_county_poolsub_GCS_v20b.csv')
naep <- filter(naep, subgroup == 'all')
naep <- select(naep, 1:4, mn_avg_mth_eb)
naep$countyid <- str_pad(naep$countyid, 5, 'left', '0')

df_gsc_mth <- select(naep, countyid, mn_avg_mth_eb) %>% filter(!is.na(mn_avg_mth_eb)) %>%
  mutate(mn_avg_mth_eb = round(mn_avg_mth_eb, 2) - 5.5) %>%
  mutate(mn_avg_mth_eb = cut(mn_avg_mth_eb,
                             breaks = c(-4,-2, -1, 0, 1, 2, 3),
                             labels = c('2 or more grade levels below',
                                        '1 to 2 grade levels below',
                                        '0 to 1 grade levels below',
                                        '0 to 1 grade levels above',
                                        '1 to 2 grade levels above',
                                        '2 or more grade levels above'
                                        )))
colnames(df_gsc_mth) <- c('GEOID','value')

us.map <- readOGR(dsn = ".", layer = "cb_2017_us_county_20m", stringsAsFactors = FALSE)
us.map <- us.map[!us.map$STATEFP %in% c("72", "66", "78", "60", "69",
                                        "64", "68", "70", "74"),]
us.map <- us.map[!us.map$STATEFP %in% c("81", "84", "86", "87", "89", "71", "76",
                                        "95", "79"),]
leafmap_mth <- merge(us.map, df_gsc_mth, by=c("GEOID"))

popup_dat <- paste0("<strong>County: </strong>", 
                    leafmap_mth$NAME, 
                    "<br><strong>Value: </strong>", 
                    leafmap_mth$value)

pal <- colorFactor("RdBu", domain = leafmap_mth$value, reverse = FALSE)
```

Using the data from Standord Educationa Data Archive [(SEDA)](https://cepa.stanford.edu/seda/overview), we can descriptively show the mathematics achievement outcomes of the US counties and how they compare among each other. 

Each county is given an estimate of the average grade level of mathematics performance pooled across all students from grade 3 to 8 and from the years of 2009 to 2016. Since only students from grade 3 to 8 are included in the data, the estiamtes are scaled so that achievement on average is 5.5 grade level. The estimates from all counties are then compared with this average to see how many grade levels are they above or below this average.

```{r echo = FALSE}
leaflet(data = leafmap_mth) %>% addTiles() %>%
  addPolygons(fillColor = ~pal(value), 
              fillOpacity = 0.8, 
              color = "#909096", 
              weight = 1,
              popup = popup_dat) %>%
  addLegend("bottomright", pal = pal, values = ~value,
            title = "Average Grade Level Compared to National Average",
            opacity = 1
  ) %>% setView(lat = 40.7529453, lng = -94.3954066, zoom = 4)
```

<p></p>

This visualization is purely exploratory in nature and no claims are made, but the patterns observed are very interesting nonetheless.

Technical information and details about how the the data were collected and how the grade levels were estimated can be found in the webite of [SEDA](https://cepa.stanford.edu/seda/overview).